Phonon polaritonics in two-dimensional materials

### 2019-03-19 paper.tex
Change referencing style to achemso
Last bits

Next, we express the polarization matrix element in Equation~\eqref{eq:polpolresponse} in terms of the Born effective charges of the monolayer and the phonon displacement eigenvectors. Considering the longitudinal phonon contribution to the second-quantized ionic displacement, as in Ref.~\citenum{srivastava1990physics}, we find that the conductivity within the RPA is given by
\begin{equation}
    \sigma(\mathbf{q},\omega) = -\frac{\iu\omega }{\Omega}\frac{\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\kappa} \Big|^2}{\omega^2_{\mathbf{q},\mathrm{L}}-\omega^2-\iu\omega\tau^{-1}}.
    \label{eq:conductivity_simplified}
\end{equation} 
We have defined scaled eigendisplacements $\boldsymbol{\eta}_{\kappa\mathbf{q}}\equiv \frac{\hat{\mathbf{e}}_{\kappa\mathbf{q},\mathrm{L}}}{\sqrt{M_{\kappa}}}$, where $\hat{\mathbf{e}}_{\kappa\mathbf{q},\mathrm{L}}$ is the unit-normalized polarization vector of atom $\kappa$ in the unit cell oscillating according to a longitudinal phonon of wavevector $\mathbf{q}$ and $M_{\kappa}$ is the mass of atom $\kappa$.

While the conductivity is the main electromagnetic quantity of interest for electrodynamics applications, we briefly state the form of the (2D) permittivity, as its zeros immediately yield the longitudinal modes of the system, which are the LO phonons. The permittivity within the RPA, denoted $\epsilon_{\textrm{RPA}}$ is related to the polarization-polarization response function via~\cite{jablan2009plasmonics} $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{1}{2}q\Pi(q,\omega)$ and the conductivity via $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{\iu q\sigma(q,\omega)}{2\omega\epsilon_0}$. Here, $\epsilon_{\mathrm{env}}$ is the average permittivity of the bulk above and below the monolayer, and is added to take into account the polarization arising from these bulk materials. Note that we have neglected any intrinsic high-frequency screening in the monolayer itself, which is only relevant for wavevectors comparable to the inverse layer spacing between monolayers. When considering non-local corrections to the conductivity at these large wavevectors, these must be taken into account~\cite{sohier2016two,thygesen2017calculating,sohier2017breakdown}. Based on Equation~\eqref{eq:conductivity_simplified}, the zeros $\omega_{\mathbf{q}}$ of the RPA dielectric function satisfy:
\begin{equation}
    \omega^2_{\mathbf{q}} - \omega^2_{\mathrm{TO}} = \frac{V(q)}{e^2}\frac{1}{\Omega}q^2\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\kappa}\Big|^2,
    \label{eq:zerosofresponse}
\end{equation}
  where $V(q)$ is the Coulomb interaction in Fourier space, which in two dimensions, is given by $V(q) = \frac{e^2}{2\epsilon_0\epsilon_{\mathrm{env}}q}$. Given that the zeros of the dielectric function are associated with longitudinal modes, one expects that $\omega_{\mathbf{q}}$ is in fact the frequency of the LO phonon mode. This is consistent with the result of Ref.~\citenum{sohier2017breakdown}, in which \relax{{}\ }\citeauthor{sohier2017breakdown} show that in 2D polar materials, the extra restoring forces on LO phonons relative to TO phonons, due to the Coulomb interaction, lead to a wavevector-dependent LO--TO splitting and zero LO--TO splitting at the $\Gamma$ point of the Brillouin zone.

@@

In the rest of this section, we discuss the relation between the dispersion of an hBN monolayer versus the atomically-thick limit of a thin film of a material with hBN's bulk dielectric function. To aid this discussion, in Figure~\ref{fig:2}(c), we show the dispersion relation of thin films of bulk hBN with film thicknesses of 1, 2, 4, and 8 times the interlayer spacing of bulk hBN, which is roughly \SI{0.33}{\nm}. For these plots, we take hBN to be cleaved such that the optical axis is perpendicular to the plane of the film. The components of the bulk permittivity perpendicular and parallel to the c-axis ($\epsilon_{\perp}$ and $\epsilon_{\parallel}$, respectively; indexed by $\alpha\in\{\perp,\parallel\}$ below) are then given by
\begin{equation}
	\epsilon_{\alpha}(\omega) = \epsilon_{\infty,\alpha}\Bigg(1 + \frac{\omega^2_{\mathrm{LO},\alpha}-\omega^2_{\mathrm{TO},\alpha}}{\omega^2_{\mathrm{TO},\alpha}-\omega^2} \Bigg),
	\label{eq:3deps}
\end{equation}
with $\epsilon_{\infty,\parallel} = \num{2.95}$, $\omega_{\mathrm{TO},\parallel} = \SI{760}{\per\cm}$, and $\omega_{\mathrm{LO},\parallel} = \SI{830}{\per\cm}$; and $\epsilon_{\infty,\perp} = \num{4.87}$, $\omega_{\mathrm{TO},\perp} = \SI{1360}{\per\cm}$, and $\omega_{\mathrm{LO},\perp} = \SI{1614}{\per\cm}$\cite{caldwell2014sub,dai2014tunable}.
Losses are ignored in this discussion altogether. 
In the range between $\omega_{\mathrm{TO}, \alpha}$ and $\omega_{\mathrm{LO}, \alpha}$, the corresponding component of the permittivity is negative, while the other component is positive. This hyperbolicity leads to a dispersion for hBN thin films that have multiple branches at a given frequency, as can be seen in Figure~\ref{fig:2}(c).  This trend persists even when the thickness of the bulk is taken down to a single layer, albeit pushed to high wavevectors. This is in contrast with the true monolayer, where there is only one LO phonon mode. Given that the phonon polariton of 2D is the LO phonon, there can be only one branch of the dispersion. 
\addition{%
The bulk dielectric function of Eq.~\eqref{eq:3deps} fails to capture this removal of higher-order modes (\ie hyperbolicity).
For a bilayer however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers, and thus one expects to see a higher-order mode. 
However, it is expected challenging to observe such a mode, as can be anticipated from the blue higher-order mode in Figure~\ref{fig:2}(c), because even for a polariton wavelength of \SI{12}{\nm}, which is very short, the polariton frequency is only a few \si{\per\cm} above the TO frequency, within the damping linewidth of the TO frequency. 
}

@@

While we have focused on anomalies between the atomically-thin limit of bulk and a true monolayer, a comparison of the monolayer with the fundamental phonon polariton mode of the one-atom-thick thin film suggests that we have a reasonable qualitative understanding of the monolayer dispersion from the one-atom-thick thin film. 
%
\addition{%
This qualitative similarity is to be somewhat expected, as a 2D layer can be considered as a very thin film of bulk material, provided that the microscopic properties of the bulk and monolayer do not deviate substantially. 
The differences between the $t = t_{\text{2D}} = \SI{0.33}{\nm}$ bulk thin-film and monolayer that nevertheless do exist, reflect such microscopic deviations.
Occasionally, \eg for compatibility with standard numerical tools, it is useful to introduce a fictitious, effective bulk material whose fundamental thin-film mode \emph{exactly} matches the monolayer's. 
The dielectric function of such a ``bulkified'' monolayer is thickness-dependent and anisotropic: for a film in the $xy$ plane, it is $\boldsymbol{\epsilon}_{\text{Q2D}} = (\hat{\mathbf{x}}\hat{\mathbf{x}} + \hat{\mathbf{y}}\hat{\mathbf{y}})\epsilon_{\text{Q2D}, \parallel} + \hat{\mathbf{z}}\hat{\mathbf{z}}\epsilon_{\text{Q2D}, \perp}$ with $\epsilon_{\text{Q2D}, \perp} = 1$ and
\begin{equation}
	\epsilon_{\text{Q2D}, \parallel}(\omega) 
	=
	 1 + \frac{\iu\sigma}{\varepsilon_0\omega t} 
 	=
	 1 + \frac{1}{t}\frac{4\epsilon_{\mathrm{env}}\omega_{\mathrm{TO}}v_\mathrm{g}}{\omega^2-\omega_{\mathrm{TO}}^2+\iu\omega\tau^{-1}}.
\end{equation}
}

@@

In Figure~\ref{fig:3}, we consider the coupling of a dipole emitter to localized phonon polaritons of nanostructured monolayer hBN. For simplicity, we consider hBN nanostructured as a disk, which leads to the formation of sharp resonances quantized along the azimuthal and radial directions. The disk is taken to have a radii varying from \SI{10}{\nm} to \SI{40}{nm}
and a loss rate $\tau^{-1} = \SI{0.5}{\meV}$, which is of the order of the loss rate in bulk hBN~\cite{caldwell2014sub, dai2014tunable}. We also show (dashed line) the results for a disk of infinite radius, i.e., a flat sheet of monolayer hBN. We parameterize the coupling between the dipole and phonon polaritons through the rate of spontaneous emission $\Gamma$ of phonon polaritons by the dipole, normalized to the rate of spontaneous emission in free space $\Gamma_0$. It is related to the dyadic Green function $\mathbf{G}$ of the Maxwell equations for the nanostructure via the expression~\cite{novotny2012principles}:
\begin{equation}
	\frac{\Gamma}{\Gamma_0} 
	=  \frac{6\pi c}{\omega}\hat{\mathbf{n}}\cdot\Im\mathbf{G}(\mathbf{r},\mathbf{r},\omega)\cdot\hat{\mathbf{n}},
\end{equation}
where $\hat{\mathbf{n}}$ is the orientation of the dipole, $\mathbf{r}$ its position, and $\omega$ its frequency.

@@

\bibliographystyle{apsrev4-1}
\bibliography{references}

>>>>

Next, we express the polarization matrix element in Equation~\eqref{eq:polpolresponse} in terms of the Born effective charges of the monolayer and the phonon displacement eigenvectors. Considering the longitudinal phonon contribution to the second-quantized ionic displacement, as in Ref.~\citenum{srivastava1990physics}, we find that the conductivity within the RPA is given by
\begin{equation}
    \sigma(\mathbf{q},\omega) = -\frac{\iu\omega }{\Omega}\frac{\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\kappa} \Big|^2}{\omega^2_{\mathbf{q},\mathrm{L}}-\omega^2-\iu\omega\tau^{-1}}.
    \label{eq:conductivity_simplified}
\end{equation} 
We have defined scaled eigendisplacements $\boldsymbol{\eta}_{\kappa\mathbf{q}}\equiv {\hat{\mathbf{e}}_{\kappa\mathbf{q},\mathrm{L}}}/{\sqrt{M_{\kappa}}}$, where $\hat{\mathbf{e}}_{\kappa\mathbf{q},\mathrm{L}}$ is the unit-normalized polarization vector of atom $\kappa$ in the unit cell oscillating according to a longitudinal phonon of wavevector $\mathbf{q}$ and $M_{\kappa}$ is the mass of atom $\kappa$.

While the conductivity is the main electromagnetic quantity of interest for electrodynamics applications, we briefly state the form of the (2D) permittivity, as its zeros immediately yield the longitudinal modes of the system, which are the LO phonons. The permittivity within the RPA, denoted $\epsilon_{\textrm{RPA}}$ is related to the polarization-polarization response function via~\cite{jablan2009plasmonics} $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + \frac{1}{2}q\Pi(q,\omega)$ and the conductivity via $\epsilon_{\textrm{RPA}} = \epsilon_{\mathrm{env}} + {\iu q\sigma(q,\omega)}/{2\epsilon_0\omega}$. Here, $\epsilon_{\mathrm{env}}$ is the average permittivity of the bulk above and below the monolayer, and is added to take into account the polarization arising from these bulk materials. Note that we have neglected any intrinsic high-frequency screening in the monolayer itself, which is only relevant for wavevectors comparable to the inverse layer spacing between monolayers. When considering non-local corrections to the conductivity at these large wavevectors, these must be taken into account~\cite{sohier2016two,thygesen2017calculating,sohier2017breakdown}. Based on Equation~\eqref{eq:conductivity_simplified}, the zeros $\omega_{\mathbf{q}}$ of the RPA dielectric function satisfy:
\begin{equation}
    \omega^2_{\mathbf{q}} - \omega^2_{\mathrm{TO}} = \frac{V(q)}{e^2}\frac{1}{\Omega}q^2\Big|\hat{\mathbf{q}}\cdot\sum\limits_{\kappa}\mathbf{Z}_{\kappa}\boldsymbol{\eta}_{\kappa}\Big|^2,
    \label{eq:zerosofresponse}
\end{equation}
  where $V(q)$ is the Coulomb interaction in Fourier space, which in two dimensions, is given by $V(q) = {e^2}/{2\epsilon_0\epsilon_{\mathrm{env}}q}$. Given that the zeros of the dielectric function are associated with longitudinal modes, one expects that $\omega_{\mathbf{q}}$ is in fact the frequency of the LO phonon mode. This is consistent with the result of Ref.~\citenum{sohier2017breakdown}, in which \relax{{}\ }\citeauthor{sohier2017breakdown} show that in 2D polar materials, the extra restoring forces on LO phonons relative to TO phonons, due to the Coulomb interaction, lead to a wavevector-dependent LO--TO splitting and zero LO--TO splitting at the $\Gamma$ point of the Brillouin zone.

@@

In the rest of this section, we discuss the relation between the dispersion of an hBN monolayer versus the atomically-thick limit of a thin film of a material with hBN's bulk dielectric function. To aid this discussion, in Figure~\ref{fig:2}(c), we show the dispersion relation of thin films of bulk hBN with film thicknesses of 1, 2, 4, and 8 times the interlayer spacing of bulk hBN, which is roughly \SI{0.33}{\nm}. For these plots, we take hBN to be cleaved such that the optical axis is perpendicular to the plane of the film. The components of the bulk permittivity perpendicular and parallel to the c-axis ($\epsilon_{\perp}$ and $\epsilon_{\parallel}$, respectively; indexed by $\alpha\in\{\perp,\parallel\}$ below) are then given by
\begin{equation}
	\epsilon_{\alpha}(\omega) = \epsilon_{\infty,\alpha}\Bigg(1 + \frac{\omega^2_{\mathrm{LO},\alpha}-\omega^2_{\mathrm{TO},\alpha}}{\omega^2_{\mathrm{TO},\alpha}-\omega^2} \Bigg),
	\label{eq:3deps}
\end{equation}
with $\epsilon_{\infty,\parallel} = \num{2.95}$, $\omega_{\mathrm{TO},\parallel} = \SI{760}{\per\cm}$, and $\omega_{\mathrm{LO},\parallel} = \SI{830}{\per\cm}$; and $\epsilon_{\infty,\perp} = \num{4.87}$, $\omega_{\mathrm{TO},\perp} = \SI{1360}{\per\cm}$, and $\omega_{\mathrm{LO},\perp} = \SI{1614}{\per\cm}$\cite{caldwell2014sub,dai2014tunable}.
Losses are ignored in this discussion altogether. 
In the range between $\omega_{\mathrm{TO}, \alpha}$ and $\omega_{\mathrm{LO}, \alpha}$, the corresponding component of the permittivity is negative, while the other component is positive. This hyperbolicity leads to a dispersion for hBN thin films that have multiple branches at a given frequency, as can be seen in Figure~\ref{fig:2}(c).  This trend persists even when the thickness of the bulk is taken down to a single layer, albeit pushed to high wavevectors. This is in contrast with the true monolayer, where there is only one LO phonon mode. Given that the phonon polariton of 2D is the LO phonon, there can be only one branch of the dispersion. 
\addition{%
The bulk dielectric function of Eq.~\eqref{eq:3deps} fails to capture this removal of higher-order modes (\ie hyperbolicity).
For a bilayer, however, there is a higher-order LO mode, associated with out-of-phase oscillation between the two layers.
Heuristically, as can be anticipated from the blue higher-order mode in Figure~\ref{fig:2}(c), this mode would be challenging to observe, given its extremely high confinement and small spectral separation from the TO mode (a few \si{\per\cm}, comparable to the damping linewidth).
}

@@

While we have focused on anomalies between the atomically-thin limit of bulk and a true monolayer, a comparison of the monolayer with the fundamental phonon polariton mode of the one-atom-thick thin film suggests that we have a reasonable qualitative understanding of the monolayer dispersion from the one-atom-thick thin film. 
%
\addition{%
This qualitative similarity is to be somewhat expected, as a 2D layer can be considered as a very thin film of bulk material, provided that the microscopic properties of the bulk and monolayer do not deviate substantially. 
The differences between the $t = t_{\text{2D}} = \SI{0.33}{\nm}$ bulk thin-film and monolayer that nevertheless do exist, reflect such microscopic deviations.
Occasionally, \eg for compatibility with standard numerical tools, it is useful to introduce a fictitious, effective bulk material whose fundamental thin-film mode \emph{exactly} matches the monolayer's. 
The dielectric function of such a ``bulkified'' monolayer is thickness-dependent and anisotropic: for a film in the $xy$ plane, it is $\boldsymbol{\epsilon}_{\text{Q2D}} = (\hat{\mathbf{x}}\hat{\mathbf{x}} + \hat{\mathbf{y}}\hat{\mathbf{y}})\epsilon_{\text{Q2D}, \parallel} + \hat{\mathbf{z}}\hat{\mathbf{z}}\epsilon_{\text{Q2D}, \perp}$ with $\epsilon_{\text{Q2D}, \perp} = 1$ and
\begin{equation}
	\epsilon_{\text{Q2D}, \parallel}(\omega) 
	=
	 1 + \frac{\iu\sigma}{\varepsilon_0\omega t} 
 	=
	 1 + \frac{1}{t}\frac{4\epsilon_{\mathrm{env}}\omega_{\mathrm{TO}}v_\mathrm{g}}{\omega_{\mathrm{TO}}^2-\omega^2-\iu\omega\tau^{-1}}.
\end{equation}
}

@@

In Figure~\ref{fig:3}, we consider the coupling of a dipole emitter to localized phonon polaritons of nanostructured monolayer hBN \addition{(we also compare extended monolayer and thin-film hBN in the Supporting Information)}.
For simplicity, we consider hBN nanostructured as a disk, which leads to the formation of sharp resonances quantized along the azimuthal and radial directions. The disk is taken to have a radii varying from \SI{10}{\nm} to \SI{40}{nm}
and a loss rate $\tau^{-1} = \SI{0.5}{\meV}$, which is of the order of the loss rate in bulk hBN~\cite{caldwell2014sub, dai2014tunable}. We also show (dashed line) the results for a disk of infinite radius, i.e., a flat sheet of monolayer hBN. We parameterize the coupling between the dipole and phonon polaritons through the rate of spontaneous emission $\Gamma$ of phonon polaritons by the dipole, normalized to the rate of spontaneous emission in free space $\Gamma_0$. It is related to the dyadic Green function $\mathbf{G}$ of the Maxwell equations for the nanostructure via the expression~\cite{novotny2012principles}:
\begin{equation}
	\frac{\Gamma}{\Gamma_0} 
	=  \frac{6\pi c}{\omega}\hat{\mathbf{n}}\cdot\Im\mathbf{G}(\mathbf{r},\mathbf{r},\omega)\cdot\hat{\mathbf{n}},
\end{equation}
where $\hat{\mathbf{n}}$ is the orientation of the dipole, $\mathbf{r}$ its position, and $\omega$ its frequency.

@@

\bibliographystyle{achemso}
\bibliography{references}
